% ===========================================================================
%  APPENDIX
% ===========================================================================
\chapter{Data Dictionary and Model Parameters}
\label{app:data_params}


% ===================================================================
\section{Complete Variable Dictionary}
\label{app:data_dict}

Table~\ref{tab:app_variables} lists all variables used across the three chapters, including source, frequency, transformation, and coverage.

\begin{table}[htbp]
\centering
\caption{Complete variable dictionary}
\label{tab:app_variables}
\small
\begin{tabular}{@{} l l l l l @{}}
\toprule
{Variable} & {Source} & {Freq.} & {Transform} & {Coverage} \\
\midrule
\multicolumn{5}{@{}l}{\emph{Target variables}} \\
\addlinespace
Current account balance & ECB SDW (BP6) & Q & Level & 2000\,Q1--2022\,Q4 \\
Goods trade balance & ECB SDW (BP6) & M & Level & 2000\,M1--2022\,M12 \\
\addlinespace
\multicolumn{5}{@{}l}{\emph{Conventional macro predictors}} \\
\addlinespace
Industrial production & ECB SDW & M & $\Delta\log$ & 2000\,M1--2022\,M12 \\
Retail sales & Eurostat & M & $\Delta\log$ & 2000\,M1--2022\,M12 \\
Unemployment rate & ECB SDW & M & $\Delta$ & 2000\,M1--2022\,M12 \\
Manufacturing PMI & Markit/S\&P & M & Level & 2000\,M1--2022\,M12 \\
Consumer confidence & Eurostat & M & Level & 2000\,M1--2022\,M12 \\
Goods exports & ECB SDW & M & $\Delta\log$ & 2000\,M1--2022\,M12 \\
Goods imports & ECB SDW & M & $\Delta\log$ & 2000\,M1--2022\,M12 \\
Terms of trade & ECB SDW & M & $\Delta\log$ & 2000\,M1--2022\,M12 \\
EUR/USD exchange rate & ECB SDW (EXR) & M & $\Delta\log$ & 2000\,M1--2022\,M12 \\
Euribor 3-month & ECB SDW (FM) & M & $\Delta$ & 2000\,M1--2022\,M12 \\
EURO STOXX 50 returns & ECB SDW & M & Returns & 2000\,M1--2022\,M12 \\
HICP inflation & ECB SDW (ICP) & M & Level & 2000\,M1--2022\,M12 \\
\addlinespace
\multicolumn{5}{@{}l}{\emph{Alternative data (Chapter~2)}} \\
\addlinespace
Baltic Dry Index & FRED & D$\to$M & $\Delta\log$ & 2000\,M1--2022\,M12 \\
Container Throughput & OECD & M & $\Delta\log$ & 2008\,M1--2022\,M12 \\
EUR/USD realised vol. & FRED & D$\to$M & Level & 2000\,M1--2022\,M12 \\
FR--DE 10Y spread & FRED & D$\to$M & Level & 2000\,M1--2022\,M12 \\
Trade Policy Uncert. & Caldara--Iacoviello & M & $\Delta\log$ & 2000\,M1--2022\,M12 \\
Google Trends (exports) & Google Trends & W$\to$M & Level & 2004\,M1--2022\,M12 \\
ECB FinBERT sentiment & ECB press conferences & M & Level & 2008\,M1--2022\,M12 \\
BdF Business Climate & FRED & M & Level + $\Delta$ & 2000\,M1--2022\,M12 \\
Nighttime light intensity & VIIRS/DMSP$^*$ & M & $\Delta\log$ & 2008\,M1--2022\,M12 \\
\bottomrule
\end{tabular}
\vspace{0.5em}

\begin{minipage}{\textwidth}
\footnotesize
$^*$ Synthetic proxy for pre-2012 data, calibrated on overlap period with industrial production.\\
D$\to$M = daily aggregated to monthly average.  W$\to$M = weekly aggregated to monthly average.\\
$\Delta\log$ = first difference of natural logarithm.  $\Delta$ = first difference.
\end{minipage}
\end{table}


% ===================================================================
\section{Model Hyperparameters}
\label{app:hyperparams}

Table~\ref{tab:app_hyperparams} reports the hyperparameter settings for all nine models evaluated in Chapter~1.

\begin{table}[htbp]
\centering
\caption{Model hyperparameters}
\label{tab:app_hyperparams}
\small
\begin{tabular}{@{} l p{10cm} @{}}
\toprule
{Model} & {Parameters} \\
\midrule
AR(1) & OLS, 1 lag. \\
\addlinespace
AR(4) & OLS, 4 lags. \\
\addlinespace
DFM & Number of factors $k$ selected automatically at each expanding-window step using the $IC_{p2}$ criterion of \citet{bai2002determining}, searching $k \in \{1,\ldots,8\}$.  Principal components extracted via PCA from standardised predictor matrix; linear forecasting equation estimated by OLS. \\
\addlinespace
Bridge & BIC-based stepwise predictor selection (forward entry from all available monthly indicators).  Quarterly target regressed on quarterly average of selected predictor(s).  OLS estimation. \\
\addlinespace
Ridge & $\ell_2$ penalty, $\alpha$ selected by efficient leave-one-out cross-validation from $\{0.01, 0.1, 1, 10, 100\}$.  Features standardised (zero mean, unit variance). \\
\addlinespace
Gradient Boosting & Hyperparameters selected by 3-fold \texttt{TimeSeriesSplit} cross-validation at each expanding-window step over a grid: \texttt{learning\_rate} $\in \{0.05, 0.1\}$, \texttt{max\_depth} $\in \{3, 5\}$, \texttt{n\_estimators} $\in \{100, 200\}$.  \texttt{min\_samples\_leaf}$= 5$. \\
\addlinespace
XGBoost & Same 3-fold TSCV protocol as GB with grid: \texttt{learning\_rate} $\in \{0.05, 0.1\}$, \texttt{max\_depth} $\in \{3, 5\}$, \texttt{reg\_lambda} $\in \{0, 1\}$.  Fixed: \texttt{n\_estimators}$= 100$, \texttt{colsample\_bytree}$= 0.8$, \texttt{min\_child\_weight}$= 5$ \citep{chen2016xgboost}. \\
\addlinespace
LSTM & 1 layer, 32 hidden units, lookback $= 4$, dropout $= 0.2$ between recurrent and output layers.  Adam optimiser (lr $= 0.005$).  Validation set (last 20\% of training window) used for early stopping (patience 15) and \texttt{ReduceLROnPlateau} scheduler (factor~0.5, patience~5).  Best validation-loss state restored.  100 epochs max.  PyTorch implementation. \\
\addlinespace
GRU & 2 layers, 32 hidden units/layer, lookback $= 4$, dropout $= 0.1$, weight decay $= 10^{-4}$.  Adam optimiser (lr $= 0.005$).  100 epochs max, early stopping (patience 15).  Replaces LSTM's 3 gates with 2 (reset, update). \\
\bottomrule
\end{tabular}
\end{table}


% ===================================================================
\section{Bootstrap Details}
\label{app:bootstrap}

\subsection{Block Bootstrap Procedure}

The block bootstrap \citep{kunsch1989jackknife} is used throughout to compute confidence intervals for RMSE ratios.  The procedure is as follows:

\begin{enumerate}
  \item Let $\{e_{1,t}, e_{2,t}\}_{t=1}^{T}$ be the forecast error sequences for two competing models over the $T$-period evaluation window.
  \item Choose block length $b$ (4 quarters for Chapter~1; 6 months for Chapter~2) to preserve temporal dependence in the error sequence.
  \item For $r = 1, \ldots, R$ ($R = 1{,}000$ replications):
    \begin{enumerate}
      \item Draw $\lceil T/b \rceil$ block starting indices uniformly from $\{1, \ldots, T - b + 1\}$.
      \item Concatenate the blocks and trim to length $T$, yielding resampled error pairs $\{e_{1,t}^{*(r)}, e_{2,t}^{*(r)}\}$.
      \item Compute $\text{RMSE}^{*(r)}_k = \sqrt{\frac{1}{T} \sum_t (e_{k,t}^{*(r)})^2}$ for $k = 1, 2$.
      \item Record the RMSE ratio $\rho^{*(r)} = \text{RMSE}^{*(r)}_2 / \text{RMSE}^{*(r)}_1 - 1$.
    \end{enumerate}
  \item Report the 2.5th and 97.5th percentiles of $\{\rho^{*(r)}\}_{r=1}^{R}$ as the 95\% confidence interval.
\end{enumerate}

\subsection{Block Length Selection}

The block length $b$ is set to match one complete seasonal cycle of the target frequency: $b = 4$ for quarterly data (one year) and $b = 6$ for monthly data (one half-year).  Sensitivity analysis with $b \in \{2, 4, 6, 8\}$ for quarterly data shows that the confidence intervals are robust to this choice: the Ridge improvement remains significant for all block lengths tested.


% ===================================================================
\section{NLP Pipeline Technical Details}
\label{app:nlp_details}

\subsection{FinBERT Configuration}

We use the \texttt{ProsusAI/finbert} model \citep{araci2019finbert} accessed through the Hugging Face \texttt{transformers} library (version 4.x).  Each text is tokenised with a maximum sequence length of 512 tokens, with truncation enabled.  The model outputs a three-class probability distribution (positive, negative, neutral); we compute a continuous sentiment score as:
\[
  s_t = P(\text{positive}) - P(\text{negative})
\]
ranging from $-1$ (fully negative) to $+1$ (fully positive).

\subsection{ECB Press Conference Processing}

ECB monetary policy press conference transcripts are sourced from the official ECB website.  For each meeting:
\begin{enumerate}
  \item The full transcript is downloaded via HTTP request.
  \item The text is split into individual sentences using rule-based tokenisation.
  \item Each sentence is scored by FinBERT.
  \item Monthly aggregates are computed: (a) mean sentiment, (b) standard deviation (dispersion).
\end{enumerate}

Our scraper successfully retrieves approximately 13\% of the 148 known ECB press conference dates (2008--2022).  For dates where scraping fails, we apply linear interpolation from the nearest available values.  This limited coverage is acknowledged as a weakness; future work should use the ECB's structured data feeds or direct PDF extraction.

\subsection{Additional Text Sources}

In addition to ECB press conferences, the NLP pipeline processes:
\begin{itemize}
  \item \textbf{ECB Financial Stability Reviews} (semi-annual, ~20 documents): Longer analytical documents providing assessment of Euro-area financial conditions.
  \item \textbf{ECB Working Papers} (abstracts only, filtered by trade/BoP keywords): Research-oriented signal that may lead policy communication.
\end{itemize}


% ===================================================================
\section{Satellite Data Construction}
\label{app:satellite}

Nighttime light intensity data from the Visible Infrared Imaging Radiometer Suite (VIIRS) Day/Night Band are available at monthly resolution from April 2012.  For the earlier period (2008--2012), we construct a synthetic proxy using the following procedure:

\begin{enumerate}
  \item For the overlap period (2012\,M4--2015\,M12), estimate the relationship:
    \[
      \text{NTL}_t = \alpha + \beta_1 \text{IP}_t + \beta_2 \text{Elec}_t + \varepsilon_t
    \]
    where $\text{IP}_t$ is industrial production and $\text{Elec}_t$ is electricity consumption.
  \item Use the fitted model to backcast $\widehat{\text{NTL}}_t$ for the pre-VIIRS period.
  \item Splice the synthetic and observed series at the overlap point.
\end{enumerate}

This is an acknowledged approximation.  The synthetic proxy captures the low-frequency trend of economic activity but cannot replicate the short-run dynamics specific to nighttime lights (e.g., construction activity, seasonal lighting patterns).


% ===================================================================
\section{Reproducibility}
\label{app:reproducibility}

\subsection{Software Environment}

All experiments were conducted using:
\begin{itemize}
  \item \textbf{Python} 3.11.9 (64-bit, Windows)
  \item \textbf{Key packages:} pandas 3.x, scikit-learn 1.x, xgboost 3.x, PyTorch 2.x (CUDA), transformers 5.x (Hugging Face), shap 0.x, statsmodels 0.x, scipy 1.x
  \item \textbf{Hardware:} NVIDIA GPU with CUDA support (for LSTM/GRU training)
\end{itemize}

\subsection{Data Access}

All data sources are publicly available without authentication:
\begin{itemize}
  \item ECB Statistical Data Warehouse: \url{https://data.ecb.europa.eu}
  \item FRED (Federal Reserve Economic Data): \url{https://fred.stlouisfed.org}
  \item Eurostat: \url{https://ec.europa.eu/eurostat}
  \item Google Trends: \url{https://trends.google.com}
  \item Caldara--Iacoviello Geopolitical Risk Index: accessed via FRED
\end{itemize}

\subsection{Code Availability}

The complete codebase, including data download scripts, model implementations, evaluation pipelines, and this LaTeX thesis, is available at:
\begin{center}
\url{https://github.com/Varcolacus/phd_bop_nowcasting}
\end{center}
